---
title: "Gapminder Wrangle"
author: "Julie"
date: "3/6/2018"
output: html_document
---

We're going to wrangle gapminder data using `dplyr`.

```{r}
library(tidyverse)  # install.packages('tidyverse')

## read in data from a url
gapminder <- read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')
```

Explore the dataframe as a whole 
```{r}

head(gapminder)
tail(gapminder)
head(gapminder, 10)

## look at structure
str(gapminder)

## some other info about our dataframe
names(gapminder)
dim(gapminder)
ncol(gapminder)
nrow(gapminder)

## create dim from ncol nrow
c( nrow(gapminder), ncol(gapminder) )


## summary statistics
summary(gapminder)

library(skimr) # install.packages('skimr')
skim(gapminder)
```


Explore inside the dataframe `$`
```{r}
gapminder$lifeExp
head(gapminder$lifeExp)
str(gapminder$lifeExp)
```

Using dplyr!
```{r}
## filter by rows
filter(gapminder, lifeExp < 29)
filter(gapminder, country == "Mexico")

## to filter 2 countries use the `%in%` operator
x <- filter(gapminder, country %in% c("Mexico", "Chile"))

## filter one country in a specific year
filter(gapminder, country == "Mexico", year == 2002)

## select by columns
select(gapminder, year, lifeExp)

## use the `-` to deselect columns
select(gapminder, -continent, -lifeExp)

## use filter and select together
gap_cambodia <- filter(gapminder, country == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp)

```


Life changing pipe operator `%>%` "and then"
```{r}
gapminder %>% head()   # equivalent to head(gapminder)

gapminder %>% head(3)  # equivalent to head(gapminder, 3)

## use filter and select together 
gap_cambodia <- filter(gapminder, country == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp)

## now let's do that with the %>% operator!
gap_cambodia <- gapminder %>% filter(country == "Cambodia")
gap_cambodia2 <- gap_cambodia %>% select(-continent, -lifeExp)

## now let's do that with the %>% operator!
gap_cambodia <- gapminder %>% 
  filter(country == "Cambodia") %>% 
  select(-continent, -lifeExp)
```

to add variables using `mutate()`
```{r}
gapminder %>%
  mutate(index = 1:nrow(gapminder))

gapminder %>%
  mutate(gdp = pop * gdpPercap)
```

Find the maximum gpdPercap of Egypt and Vietnam, and save it in a new column. 
```{r}
## find the max of both Egypt and Vietnam together
x <- gapminder %>%
  filter(country %in% c("Egypt", "Vietnam")) %>%
  mutate(max_gdp = max(gdpPercap))

## find the max of Egypt and Vietnam separately
x2 <- gapminder %>%
  filter(country %in% c("Egypt", "Vietnam")) %>%
  group_by(country) %>%
  mutate(max_gdp = max(gdpPercap)) %>%
  ungroup() # important to remember to ungroup


str(x2)

```


summarize and group_by together!
```{r}
gapminder %>%
  group_by(country) %>%
  mutate(gdp = pop * gdpPercap) %>%
  summarize(max_gdp = max(gdp)) %>%
  ungroup() %>%
  arrange(desc(max_gdp))
```

Let's find the max lifeExp for countries in Asia
Hint max() 
```{r}
asia <- gapminder %>%
  filter(continent == "Asia") %>%
  group_by(country) %>%
  mutate(gdp = pop * gdpPercap) %>%
  summarize(max_gdp = max(gdp)) %>%
  ungroup() %>%
  arrange(desc(max_gdp))
```

Joining!
```{r}
# read in CO2 data
co2 <- read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv')

## explore
co2 %>% 
  head()

## gap 2007
gap_2007 <- gapminder %>%
  filter(year == 2007)

## left_join gap_2007 to co2
lj <- left_join(gap_2007, co2, by = "country")

## explore lj
lj %>% dim()
lj %>% summary()


## right_join gap_2007 and co2
rj <- right_join(gap_2007, co2, by = "country")


```




